---
title: "Drug and Alcohol Health Analysis"
author: "Deric Liang and David Chen"
date: "04/30/2019"
output: html_document
---

```{r message = FALSE, include = FALSE}
# Clean up R environment
rm(list = ls())

# libraries
library(data.table)
library(tidyverse)
library(lubridate)
library(leaps)
library(rpart)
library(partykit)
library(caret)
#library(sparklyr)

# load data
educ_raw <-
  fread("Data/Education.csv")
healthexp_raw <-
  fread("Data/HealthExpend.csv", header = TRUE)
# pop_raw <-
#   fread("Data/PopulationEstimates.csv")
### population counts included in drug mortality data
poverty_raw <-
  fread("Data/PovertyEstimates.csv")
unemp_raw <-
  fread("Data/Unemployment.csv")
drugmort_raw <-
  fread("Data/Compressed Mortality, 1999-2016.txt")

unint_death <- 
  fread('Data/X4045_Mortality.txt')
int_death <-
  fread('Data/X6065_Mortality.txt')

# define functions

identify_region <- function(state) {
  # Purpose: This functions identifies the region of a state
  # Inputs:
  ### state: United States state in a character format
  # Output: Which region of the United States the state applies to. If none, NA.
  
  Northeast <- c('Maine', 'New Hampshire', 'Vermont', 'Massachusetts', 
               'Rhode Island', 'Connecticut', 'New York', 'New Jersey',
               'Pennsylvania')

  Midwest <- c('Ohio', 'Michigan', 'Indiana', 'Wisconsin', 'Illinois', 
               'Minnesota', 'Iowa', 'Missouri', 'North Dakota', 
               'South Dakota', 'Nebraska', 'Kansas')
  
  South <- c('Delaware', 'Maryland', 'Virginia', 'West Virginia', 'Kentucky', 
             'North Carolina', 'South Carolina', 'Tennessee', 'Georgia', 
             'Florida', 'Alabama', 'Mississippi', 'Arkansas', 'Louisiana', 
             'Texas', 'Oklahoma')
  
  West <- c('Montana', 'Idaho', 'Wyoming', 'Colorado', 'New Mexico', 
            'Arizona', 'Utah', 'Nevada', 'California', 'Oregon', 
            'Washington', 'Alaska', 'Hawaii')
  
  if (state %in% Northeast) {
    return('Northeast')
  }
  else if (state %in% Midwest) {
    return('Midwest')
  }
  else if (state %in% South) {
    return('South') 
  }
  else if (state %in% West) {
    return('West')
  }
  else {
    return(NA)
  }
}

```

## Data Wrangling
```{r}
# Drug Mortality Data
### Remove unnecessary columns and `Total` row
drugmort <-
  drugmort_raw %>%
  na.omit() %>%
  select(-Notes, -`State Code`) %>%
  mutate(Deaths = as.numeric(Deaths)) %>%
  rename(DrugDeaths = Deaths,
         DrugMortalityRate = `Crude Rate`)

# Education Level Data
### Clean headers, select for state data, select relevant columns
educ <-
  educ_raw %>%
  distinct(State, .keep_all = TRUE) %>%
  filter(`FIPS Code` != is.na(`FIPS Code`),
         State != "PR") %>%
  select(`Area name`, `Percent of adults with a bachelor's degree or higher, 2013-17`,
         `Percent of adults completing some college or associate's degree, 2013-17`,
         `Percent of adults with a high school diploma only, 2013-17`,
         `Percent of adults with less than a high school diploma, 2013-17`) %>%
  rename(State = `Area name`,
         PercentBachelor = `Percent of adults with a bachelor's degree or higher, 2013-17`,
         PercentAssoc = `Percent of adults completing some college or associate's degree, 2013-17`,
         PercentHS = `Percent of adults with a high school diploma only, 2013-17`,
         PercentLessHS = `Percent of adults with less than a high school diploma, 2013-17`)

# Health Expenditure Data
### Remove non-state rows, extraneous column
healthexp <-
  healthexp_raw[1:51] %>%
  rename(HealthExpenditure = `2016`, State = STATE) %>%
  mutate(HealthExpenditure = parse_number(HealthExpenditure)) %>%
  select(State, HealthExpenditure)

# Poverty Data
### Clean headers, select for state data, select relevant columns
poverty_raw$State[poverty_raw$State == "Wi"] <- "WI"
poverty <-
  poverty_raw %>%
  distinct(State, .keep_all = TRUE) %>%
  filter(FIPStxt != is.na(FIPStxt)) %>%
  select(Area_Name, POVALL_2017, PCTPOVALL_2017, POV017_2017, PCTPOV017_2017) %>%
  rename(State = Area_Name,
         PovertyCountAll = POVALL_2017,
         PovertyPctAll = PCTPOVALL_2017,
         PovertyCount0_17 = POV017_2017,
         PovertyPctAll0_17 = PCTPOV017_2017) %>%
  mutate(PovertyCountAll = parse_number(PovertyCountAll),
         PovertyCount0_17 = parse_number(PovertyCount0_17))

# Unemployment Data
### Clean headers, select for state data, select relevant columns
unemp_raw$State[unemp_raw$State == "Co"] <- "CO"
unemployment <-
  unemp_raw %>%
  distinct(State, .keep_all = TRUE) %>%
  filter(FIPStxt != is.na(FIPStxt),
         State != "PR") %>%
  select(Area_name, Civilian_labor_force_2016, Unemployed_2016, Unemployment_rate_2016) %>%
  rename(State = Area_name,
         LaborForce = Civilian_labor_force_2016,
         Unemployed = Unemployed_2016,
         UnemploymentRate = Unemployment_rate_2016) %>%
  mutate(LaborForce = parse_number(LaborForce),
         Unemployed = parse_number(Unemployed))

```

```{r}

intdeath <-
  int_death %>%
  filter(State != "District of Columbia") %>%
  select(-Notes, -`State Code`) %>%
  mutate(Deaths = as.numeric(Deaths)) %>%
  rename(int_DrugDeaths = Deaths,
         int_DrugMortalityRate = `Crude Rate`) %>%
  mutate(int_DrugMortalityRate = parse_number(int_DrugMortalityRate))

unintdeath <-
  unint_death %>%
  filter(State != "District of Columbia") %>%
  select(-Notes, -`State Code`, -`Population`) %>%
  mutate(Deaths = as.numeric(Deaths)) %>%
  rename(unint_DrugDeaths = Deaths,
         unint_DrugMortalityRate = `Crude Rate`) %>%
  mutate(unint_DrugMortalityRate = parse_number(unint_DrugMortalityRate))


combined_deaths <- 
  left_join(intdeath, unintdeath, by = 'State')


HealthData <-
  left_join(combined_deaths, drugmort, by = c("State", 'Population')) %>%
  left_join(healthexp, by = "State") %>%
  left_join(educ, by = 'State') %>%
  left_join(poverty, by = 'State') %>%
  left_join(unemployment, by = 'State') %>%
  filter(State != "District of Columbia")
```



```{r}
#EDA 

ggplot(combined_deaths, aes(x = unint_DrugDeaths, y = int_DrugDeaths)) +
  geom_point()
ggplot(combined_deaths, aes(x = unint_DrugMortalityRate, y = int_DrugMortalityRate)) +
  geom_point()
```

### Data Wrangling - Regions according to the US Census Bureau:  

Sourced from : https://www.businessinsider.com/united-states-regions-new-england-midwest-south-2018-4    

1. "The Northeast includes Maine, New Hampshire, Vermont, Massachusetts, Rhode Island, Connecticut, New York, New Jersey, and Pennsylvania."  

2. "[The Midwest] consists of Ohio, Michigan, Indiana, Wisconsin, Illinois, Minnesota, Iowa, Missouri, North Dakota, South Dakota, Nebraska, and Kansas."  

3. "...the South consists of Delaware, Maryland, Virginia, West Virginia, Kentucky, North Carolina, South Carolina, Tennessee, Georgia, Florida, Alabama, Mississippi, Arkansas, Louisiana, Texas, and Oklahoma."   

4. "The West consists of Montana, Idaho, Wyoming, Colorado, New Mexico, Arizona, Utah, Nevada, California, Oregon, Washington, Alaska, and Hawaii."  

```{r}
# Attaching region to dataset
# Only concern is NA

HealthData_region <- 
  HealthData %>%
  
  # This line identifies the region and creates the variable
  mutate(region = sapply(HealthData$State, FUN = identify_region)) %>%
  
  # Create a copy of region to unnest and spread
  mutate(region_copy = region) %>%
  
  # The following code creates dummy variables for region
  unnest(region_copy) %>% 
  mutate(new = 1) %>% 
  spread(region_copy, new, fill = 0) 

```

```{r}

Mortality_asc <-
  HealthData_region %>%
  arrange(desc(DrugMortalityRate))

ggplot(Mortality_asc, aes(x = reorder(State, -DrugMortalityRate), 
                          y = DrugMortalityRate, fill = region)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(y = 'Drug Mortality Rate per 10000')


<<<<<<< HEAD
Interesting_variables <- 
  HealthData_region %>%
  select(DrugMortalityRate, HealthExpenditure, PercentHS, PovertyPctAll, 
         UnemploymentRate)

plot(Interesting_variables)
# test
=======

>>>>>>> 14a026d2e555227934eb412e05a1e13118a7b154
```


## Regression Models/Supervised Learning

```{r}
# set RNG seed for reproducible results
set.seed(538) 
# partition the data into training & test data
n <- nrow(HealthData_region)
test_idx <- sample.int(n, size = round(0.2 * n)) # select row numbers for the test set
train <- HealthData_region[-test_idx, ]  # exclude the test set cases
test <- HealthData_region[test_idx, ]

form <- 
  DrugMortalityRate ~ HealthExpenditure + PercentBachelor + PovertyPctAll + UnemploymentRate
```

### Null Model

```{r}
null_mod <-
  lm(DrugMortalityRate ~ 1, data = train)
summary(null_mod)
nullrsq <-
  1 - sum((null_mod$residuals)^2) / sum((train$DrugMortalityRate - mean(train$DrugMortalityRate))^2)
nullrsq
```
The mean drug mortality rate for the training data is 20.124, with an R-squared of 0.

### Regression

```{r}
fullhealth_lm <-
  lm(form, data = HealthData_region)
summary(fullhealth_lm)
plot(fullhealth_lm)
# Fitted model is not good, none of the regressors are significant
```

```{r}
subsets <-
  regsubsets(form, data = HealthData_region, nbest = 5)
plot(subsets)
# Model with just `UnemploymentRate` regressor the best
```

```{r}
HealthData_region %>%
  ggplot(aes(x = UnemploymentRate, y = DrugMortalityRate)) +
  geom_point(aes(color = region)) +
  stat_smooth(color = "grey50", method = lm, se = 0) +
  xlab("Unemployment Rate") +
  ylab("Drug Mortality Rate") +
  labs(title = "Drug Mortality Rate vs. Unemployment Rate")
```
A visual of the fit of Drug Mortality Rate vs. Unemployment Rate.


```{r}
final_lm <-
  lm(DrugMortalityRate ~ UnemploymentRate, data = HealthData_region)
summary(final_lm)
plot(final_lm)
```

Our regression modeling shows us that only the Unemployment Rate has a significant effect on the Drug Mortality Rate. Checking diagnostics plots, we are fine to assume linearity, independence, normality, and equivariance.

### Supervised Learning

We want to explore other modeling methods which could help us predict Drug Mortality Rate, beyond classical regression models.

```{r}
regtree <-
  rpart(form, data = train)
plot(as.party(regtree))
```

The regression tree shows that the education level (the percent of people who have a Bachelor's degree) is also significant in predicting Drug Mortality Rate, in addition to Unemployment Rate. An unemployment rate greater than or equal to 4.9% shows higher drug mortality, then the next highest shows an unemployment rate of less than 4.9% and greater than or equal to 33.55% of people with a Bachelors, then finally an unemployment rate of less than 4.9% and less than 33.55% of people with a Bachelors.

```{r}
# define training control
train_control <- 
  trainControl(method = "cv", number = 2)
# train the model
treecv <- 
  form %>%
  train(data = train, trControl = train_control, method = "rpart")
# summarize results
treecv
```

## Unsupervised Learning



## Big Data Tools









